---
title: Property Assesssment
author: RJ Cass
execute:
  echo: false
format: 
  pdf:
    include-in-header:
      text: |
        \usepackage{caption}
        \captionsetup{textfont={it}, font=small} 
        \usepackage{amsmath}
geometry: "margin=1in"
---

# Abstract

We have a dataset of sales price for homes in Iron County, with corresponding home metrics. We used this dataset to create a linear model to provide estimated home values of a given home using the following variables: lot size, square-footage of the home, number of bathrooms, whether there is a garage, whehter there is a bathroom, roof quality, and effective age. This model can also be used to provide explanations for why a given home has a certain estimated value. 

```{r read_data, include = FALSE}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(MASS))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(interactions))

HomeSales <- read_csv("https://grimshawville.byu.edu/BYUStat535/ironco.csv")
HomeSales$basmt <- as.factor(HomeSales$basmt)
HomeSales$gar <- as.factor(HomeSales$gar)
```


# Introduction

We have a dataset of the sale prices of homes in Iron County, Utah from 1990 to 1994. \textcolor{green}{Iron county is rural area, with residency consisting primarily of single-family homes. During this period in Utah, there was a significant surge in home prices (starting more or less in the early 90's)}. These data were provided by the county, where an assessor went to each property and provided various metrics about the property/home. We want to use these data to provide estimates of how much a home is worth and why, and to help home owners know what might be done to improve the value of a home. \textcolor{green}{Due to these data only coming from houses that were actually sold during that period, it's possible there are external factors that influenced which houses were sold during this time, and that this sample may not be representative of the value of all houses at the time. This could result in predictions that are inaccurate for houses outside of the dataset}. 

### Data

The dataset consists of home metrics: sale price, various 'count' metrics (bathrooms 'baths', floors 'floors', garage 'gar', basement 'basmt'), various 'quality' metrics (construction quality 'const', roof condition 'roof', home condition 'quality'), area metrics (lot acreage 'lot', home square-feet area 'area'), and age metrics (year built 'yr_built', effective age 'eff_age'). A summary of the continuous variates is provided in @tbl-summary_continuous, \textcolor{green}{and the counts of values for the discrete variates is provided in} @tbl-summary_discrete.

The quality metrics were determined by the assesor upon inspection of the property. The effective age of the home takes is based on the actual age of the home, but takes into account renovations, additions, upgrades, etc. performed on the house/property which have 'modernized' the property. Thus, it is possible to have a house built 70 years ago with an effective age of 20, and another house built 30 years ago with an effective age of 25. 

```{r tbl-summary_continuous}
#| tbl-cap: "Summary statistics of the continuous home salesprice variates"
mySummary <- function(x) {
  c(mean = round(mean(x[x != 0]), 2), sd = round(sd(x[x != 0]), 2), min = min(x[x != 0]), max = max(x[x != 0]))
}

reg_summ <- t(apply(HomeSales[, c(1:4, 6, 8:12)], 2, mySummary))
reg_var_summ <- as.data.frame(cbind(colnames(HomeSales[, c(1:4, 6, 8:12)]), reg_summ))
colnames(reg_var_summ) <- c('Variable', 'Mean', 'SD', 'Min', 'Max')
rownames(reg_var_summ) <- NULL

kable(reg_var_summ, longtable = TRUE, row.names = FALSE) %>%
  kable_styling(latex_options = "repeat_header")
```


```{r tbl-summary_discrete}
#| tbl-cap: "Counts of values of the discrete home salesprice variates"
mySummary <- function(x) {
  c(length(x[x == 1]), length(x[x == 1.5]), length(x[x == 2]), length(x[x == 3]), length(x[x == 4]), length(x[x == 5]))
}

reg_summ <- t(apply(HomeSales[, c(4:10)], 2, mySummary))
reg_var_summ <- as.data.frame(cbind(colnames(HomeSales[, c(4:10)]), reg_summ))
colnames(reg_var_summ) <- c('Variable', '1', '1.5', '2', '3', '4', '5')
rownames(reg_var_summ) <- NULL

kable(reg_var_summ, longtable = TRUE, row.names = FALSE) %>%
  kable_styling(latex_options = "repeat_header")
```

There are also 2 factor variables (garage and basement), both of which are just 0/1 indicators of whether the house has it. An important thing to note is that for the quality metrics, we have a restricted range of provided values. While the quality scale can range from 1 to 5, we only have provided values of 2 through 4 for construction quality, 2 and 3 for roof quality, and 3 and 4 for home quality. This will result in any model

### Data Exploration

\textcolor{green}{In general, using intution to explore how each metric should affect the price of a home, we expect most variables to have a positive correlation with sale price. For example, as the area of a home increases, we would expect the price to also increase. An exception to this is effective age, which we would expect to be negatively correlated (as the effective age increases, the home value would decrease).} This intution is confirmed in @fig-marginals via marginal effect plots of the continous variables. 

```{r fig-marginals, fig.height=6, fig.width=7}
#| fig-cap: "Plots showing correlation between different variates. Note the negative slope in the marginal effect of Effective Age (d)"
par(mfrow = c(2, 2))
plot(HomeSales$lot, HomeSales$price, xlab = 'Lot Size', ylab = 'Price', main = '(a) Marginal Effect of Lot Size on Price')
plot(HomeSales$area, HomeSales$price, xlab = 'House Area', ylab = 'Price', main = '(b) Marginal Effect of House Area on Price')
plot(HomeSales$yr_built, HomeSales$price, xlab = 'Year Built', ylab = 'Price', main = '(c) Marginal Effect of Year Built on Price')
plot(HomeSales$eff_age, HomeSales$price, xlab = 'Effective Age', ylab = 'Price', main = '(d) Marginal Effect of Effective Age on Price')
par(mfrow = c(1, 1))
```


\textcolor{green}{Building off of further intuition, there are certain variates that we would expect to be collinear. An example is effective age and year built: we would generally expect that a house built longer ago is also going to have a higher effective age, and thus the fields are mostly redunant. Collinear effects between variates can result in estimated coefficients that have the wrong sign and inflated standard errors. As such we will only only be including Effective Age in our model. We could also expect there to be some form of interaction between variables. For example, the difference between 1 and 2 bathrooms may be greater in a house with larger area than a house with smaller area (in a smaller house, we might consider that bathroom taking up space that may occupy some other use, and not being as valuable). As such, we will want to ensure our model explores interaction effects between the variates.}

# Methodology

With our model motivations (be able to predict a home's value, and explain to someone why it has that estimated value), we need a model that has interpretable coefficients, and that is able to make predictions on new data. 

To this end we will create a linear model of the form $y_i = x_i\beta_i + \epsilon_i, \epsilon_i \sim N(0, \sigma^2)$, where $y_i$ is the price of the home, $x_i\beta_i$ is the variates and their corresponding coefficient estimates, and $\epsilon_i$ is the vector of random noise associated with the measurements. \textcolor{green}{We will begin with a model including all variates (including all 2-way interactions, but excluding year built) and use AIC hybrid variable selection to create a model using just the most significant variables.}

This linear model will allow us to have easily interpretable continuous coefficients \textcolor{green}{(ie. holding all else constant, increasing the house quality by X amount will increase the estimated average home price by Y dollars). The discrete coefficients can be similarly interpreted, where for example, holding all else constant, a house having a garage increases the estimated average home price by Y dollars when compared to houses without a garage. These interpretations will change if the variates have any interaction effects.} Reducing the number of variables via AIC selection will further increase the interpretability of the model and its results. 

# Results

```{r model, include = FALSE}
mod <- lm(price ~ . - yr_built + lot*area + area*eff_age, data = HomeSales)
mod <- stepAIC(mod, direction = "both")
is_rmse <- format(sqrt(mean(mod$residuals^2)), scientific = FALSE)
summary(mod)
```

```{r loocv}
n <- nrow(HomeSales)
loocv_error <- numeric(n)
for (i in 1:n) {
  train <- HomeSales[-i, ]
  test <- HomeSales[i, ]

  loocv_mod <- lm(price ~  + lot*area + baths + gar + basmt + roof + eff_age, data = train)
  p <- predict(loocv_mod, newdata = test)
  loocv_error[[i]] <- (p - test$price)
}

oos_rmse <- format(round(sqrt(sum(loocv_error^2)/n), 2), scientific = FALSE)

mean_pred_error <- format(round(mean(abs(mod$residuals)), 2), scientific = FALSE)
mean_pred_perc_error <- round(mean(abs(mod$residuals / (mod$residuals - mod$fitted.values)))*100, 2)
```

We fit a model following the above outline, and the resulting variables and their estimated coefficients (and their corresponding 95% confidence intervals) are provided in @tbl-sig_coefs. As shown in the table, the selected variables are lot size, square-footage of the home, number of bathrooms, whether there is a garage, whether there is a basement, roof quality, effective age, and the interaction between lot and area. This model has an In-Sample RMSE of `r is_rmse`. We also performed Leave-one-out Cross Validation and obtained an Out of Sample RMSE of `r oos_rmse`. \textcolor{green}{These values indicate that we are not overtly overfitting, and aren't inherently trading great in-sample performance for poor out-of-sample perforamnce, or vise-versa.}

\textcolor{green}{Considering the RMSE values, these indicate that for future predictions, we can expect an uncertainty on the order of 10k dollars. When providing predictions, we will need to ensure to provide the confidence interval for the prediction. Examinig the residuals from the model even futher, it has a mean absolute prediction error of `r mean_pred_error`, and a mean absolute prediction percent error of `r mean_pred_perc_error`\%. These provide further indications of the kind of uncertainty we can expect from future predictions.}

```{r tbl-sig_coefs}
#| tbl-cap: "Estimated coefficients of final model, with their corresponding 95% Confidence Intervals"
get_sig_coefs <- function(mod) {
  coefs <- as.data.frame(summary(mod)$coefficients)
  coefs$Variable <- rownames(summary(mod)$coefficients)
  conf <- as.data.frame(confint(mod))
  conf$Variable <- rownames(confint(mod))
  coefs$Lower <- conf$`2.5 %`
  coefs$Upper <- conf$`97.5 %`
  sig_terms <- coefs[c("Variable", "Estimate", "Lower", "Upper")]
  sig_terms[, 2:4] <- round(sig_terms[, 2:4], 2)
  sig_terms
}

sig_coefs <- get_sig_coefs(mod)
# Reorder things to make lot/area next to the interaction
sig_coefs$Variable[5] <- 'Garage - Yes'
sig_coefs$Variable[6] <- 'Basement - Yes'
sig_coefs <- sig_coefs[c(1, 4:8, 2, 3, 9), ]

kable(sig_coefs, longtable = TRUE, row.names = FALSE) %>%
  kable_styling(latex_options = "repeat_header")
```

To understand the effect of Lot and Area on Price, consider @fig-int_plot: as Lot increases, the slope of the effect of Area also increases. This plot shows that even though the individual coefficients for Lot and Area are negative, the interaction dominates the effect within the range of the data. If we imagine a situation where two homeowners have the same lot sizes but different home areas, if the lot is small, the difference in estimated value will be realtively small. If the lots are large, the difference in estimated values will be larger. 

```{r fig-int_plot}
#| fig-cap: "Plot of the interaction effect of Lot and Area. As Lot increases, the slope of the Area effect increases"
lot <- seq(.05, .6, length = 5)
area <- seq(600, 2000, length = 5)
baths <- 2
gar <- 0
floors <- 1
basmt <- 1
roof <- 3
eff_age <- 5
grid <- expand.grid(lot, area, baths, gar, floors, basmt, roof, eff_age)
grid <- as.data.frame(grid)
colnames(grid) <- c('lot', 'area', 'baths', 'gar', 'floors', 'basmt', 'roof', 'eff_age')
grid$basmt <- as.factor(grid$basmt)
grid$gar <- as.factor(grid$gar)
# levels(grid$gar)
# levels(grid$basmt)
levels(grid$basmt) <- levels(HomeSales$basmt)
levels(grid$gar) <- levels(HomeSales$gar)

grid$pred <- predict(mod, newdata = grid)



ggplot(grid, aes(x = area, y = pred, color = lot, group = lot)) +
  geom_line() +
  scale_color_viridis_c() + 
  theme_minimal() +
  labs(title = "Interaction of Lot and Area",
       y = "Price", x = "Area",
       color = "Lot")
```

We can use the provided model coefficients to explain to a home-owner why their estimated home value is a certain number. For example, if they are comparing their home to their neighbor's home, we can look at the differences to explain what the difference in cost is. If their owner has a home that is the same in all ways to the neighbor's home except the owner's home has 2 bathrooms and the neighbor has 3 bathrooms, we can explain that having 1 less bathroom decreases the estimated average home price by ~\$7500 (\pm ~$4500 95% confidence interval). 

In a similar way, we can use the coefficients to inform homeowners what changes to a home would most influence their estimate home cost. If a homeowner is considering whether they should add a garage, or fix their roof, we can refer to the coefficients and show them that adding a garage would add ~\$7,600 (\pm ~$6,500 95% confidence interval) to the estimated average home's value, whereas fixing the roof enough to increase it by 1 point would increase the average estimated home value by ~\$14,900 (\pm ~$9,700 95% confidence interval). The homeowner can then look at the cost of each task to see which would be most valuable to them. 

```{r predict}
case <- data.frame(lot = 0.25, area = 1300, baths = 2, gar = factor(1, levels = c(0, 1)), 
                    basmt = factor(0, levels = c(0, 1)), roof = 3, eff_age = 5)
pred_ci <- round(predict(mod, newdata = case, , interval = "prediction", level = 0.95), 2)
```

Using this model, for a given house, we can provide an estimated range of the home's value. For example, for a house that has the parameters provided in @tbl-example_params, we predict the home will have a value within the range of `r scales::dollar(pred_ci[, 2])` to `r scales::dollar(pred_ci[, 3])`

|Variable | Value |
|:-------|:----:|
| Lot | 0.25|
|Area | 1300 |
| Baths | 2|
|Garage | Yes| 
|Basement | No | 
|Roof Quality| 3|
|Effective Age | 5|
: Example home metrics values {#tbl-example_params}


# Conclusion

We used the Iron County home sale price dataset to create a linear model. We performed variable selection to create a model that depends on lot size, square-footage of the home, number of bathrooms, whether there is a garage, whether there is a bathroom, roof quality, and effective age, with an In-Sample RMSE of `r is_rmse`. We demonstrated how this model can be used to answer expected questions from homeowners, such as why their estimated homevalue is what it is, or what added features to their house would most increase the value. 

Moving forward, we believe there are improvements that could be made to the dataset to allow for better prediction. Primarily, we believe there are factors that impact a home's value which are not included here. Some examples include a measure of how near the closest neighbor is, or how busy the road they live on is, or how close they are to a city center, etc. \textcolor{green}{We also believe this is a study that could be applied to other counties as well. If we were to include different counties, it's likely that applying the model generated on this dataset would not provide very accurate predictions due to differences in the counties (things like location, access to certain resources/features, etc.). Performing exploration on the impact of county on the home prices (ie. treating county as a factor) could detect whether there are significant differences in variate behavior across counties. }

\textcolor{green}{There are a few data points which could be outliers or overly-influential. The most clear ones are observation 53 (which has a sale price much higher than we would expect), observation 94 (the only home that has 5 baths), and the 2 observations which 4 baths (one of which has a large area which is expected, and the other which has a considerably smaller area and may be influential). In future studies, we could perform sensitivity analysis to identify whether those points have a strong impact on the model. Obtaining more data from the same period may help 'fill in the gaps' between these data points and the others, but operating under the assumption that this is all the data available, analyzing, and potentially removing, those data points could have significant impact on the model.}